The Concurrent Calculi Formalisation Benchmark

### 2023-10-05 paper/paper.tex
Add more introduction.

The authors of both POPLMark and POPLMark Reloaded
were careful to note that their benchmarks were only a beginning, and
they specifically mention reasoning about concurrency using
coinduction and linear environments as points for future work. Papers
in concurrency theory do often include mechanised proofs
(e.g.~\cite{DBLP:conf/pldi/Castro-Perez0GY21,DBLP:conf/tacas/CastroFY20,lmcs:9985,
  DBLP:journals/jar/CruzFilipeMP23, Tirore:2023}), and others also
directly discuss formalisation techniques
(e.g.~\cite{DBLP:journals/jar/BengtsonPW16, DBLP:conf/tphol/Gay01,
  DBLP:conf/ppdp/Thiemann19, DBLP:conf/forte/ZalakainD21}). However,
our experience is that choosing techniques and tools remains a
challenge with few guidelines. We propose a new collection of
benchmark problems specifically designed to tackle the issues
encountered when mechanising message-passing concurrency using process
calculi. A goal of our collection of benchmark problems is to push for
the development of guidelines and tutorials, and our challenge
problems are a simple setting in which to demonstrate and compare
techniques. Another goal is to strengthen the culture of mechanisation
by rallying the community to collaborate on exploring and developing
new tools and techniques. We have identified three key aspects which
cause issues when mechanising concurrency theory: linearity, scope
extrusion in name passing systems and infinite behaviour modelled
coinductively. Each part of our benchmark exercises one of the three aspects
independently to enable comparison and to keep the challenge problems
small. The three aspects are of course not the only ones that may
cause issues in mechanisations, but they are fundamental to the field
and will thus appear in most mechanisations of concurrency theory.
Mechanisations of actual research will additionally often require more than just
one aspect, but we do not address the combination of techniques. In
the rest of this paper we discuss the design of the challenge and
briefly describe the three challenge problems.

@@

\section{The challenge problems}
In this section we will briefly describe our three challenge problems, referring to the appendix for the full text of each challenge.
Our \textbf{linearity challenge} (Appendix~\ref{sec:challenge:linearity-beh-types}) is to prove type preservation (also known as subject reduction) for a process calculus with a session type system.
Session type systems require linear handling of typing contexts, and the proof involves reasoning about the linearity of context splitting operations.
In name passing systems, scope extrusion is the expansion of the scope of a name when the name is sent outside of its original scope.
Our \textbf{scope extrusion challenge} (Appendix~\ref{sec:challenge:name-passing-scope-extrusion}) is to prove that barbed bisimulation is an equivalence relation for an untyped name passing calculus without infinite behaviour, but with non-deterministic choice.
The proof involves reasoning about names that are ``in the process'' of being scope extruded, which presents issues for some approaches to mechanising binders.
Our \textbf{coinduction challenge} (Appendix~\ref{sec:challenge:coinduction}) is to prove that strong barbed bisimulation is an equivalence relation \footnote{AM: this seems wrong, see:  \( \sbcong{P}{Q} \) if and only if for any process \( R \) and substitution \( \sigma \), \( \sbbisim{\Ppar{\applysubst{\sigma}{P}}{R}}{\Ppar{\applysubst{\sigma}{Q}}{R}} \).}for an untyped calculus of communicating systems with (infinite) replication of processes.
The proof involves reasoning coinductively about the infinite behaviours of the replication operator.
We invite anyone interested to send us solutions to the challenge problems and to publish experience reports detailing the advantages and disadvantages of their approaches.
The interested reader can find more information about the current status of the challenge at:
\begin{center}
  \url{https://concurrentbenchmark.github.io/}
\end{center}

>>>>

The authors of both POPLMark and POPLMark Reloaded
were careful to note that their benchmarks were only a beginning, and
they specifically mention reasoning about concurrency using
coinduction and linear environments as points for future work. Papers
in concurrency theory do often include mechanised proofs
(e.g.~\cite{DBLP:conf/pldi/Castro-Perez0GY21,DBLP:conf/tacas/CastroFY20,lmcs:9985,
  DBLP:journals/jar/CruzFilipeMP23, Tirore:2023}), and others also
directly discuss formalisation techniques
(e.g.~\cite{DBLP:journals/jar/BengtsonPW16, DBLP:conf/tphol/Gay01,
  DBLP:conf/ppdp/Thiemann19, DBLP:conf/forte/ZalakainD21}). However,
our experience is that choosing techniques and tools remains a
challenge with few guidelines. We propose a new collection of
benchmark problems specifically designed to tackle the issues
encountered when mechanising message-passing concurrency using process
calculi. A goal of our collection of benchmark problems is to push for
the development of guidelines and tutorials, and our challenge
problems are a simple setting in which to demonstrate and compare
techniques. Another goal is to strengthen the culture of mechanisation
by rallying the community to collaborate on exploring and developing
new tools and techniques. We have identified three key aspects which
cause issues when mechanising concurrency theory: linearity, scope
extrusion in name passing systems and infinite behaviour modelled
coinductively. Each part of our benchmark exercises one of the three aspects
independently to enable comparison and to keep the challenge problems
small. The three aspects are of course not the only ones that may
cause issues in mechanisations, but they are fundamental to the field
and will thus appear in most mechanisations of concurrency theory.
Mechanisations of actual research will additionally often require more than just
one aspect, but we do not address the combination of techniques.

Large scale formalisation of concurrency theory is already possible.
(Paragraphs about related work in this vein)

However, it is our experience that tools and techniques are chosen in a very ad hoc fashion.
Like previous benchmark authors, we thus seek to answer a number of questions:
\begin{enumerate}
\item What is the current state of the art in mechanising process calculi?
\item What techniques and best practices can be recommended when starting formalisation projects within concurrency theory?
\item What improvements are needed to make mechanisation tools more user-friendly with regards to the issues faced when mechanising process calculi?
\end{enumerate}

Our challenge problems are designed to stimulate progress on answering these questions by providing simple environments in which to demonstrate and evaluate the many tools and techniques available.
Our problems are drawn from the basic metatheory of process calculi and include as little syntax and as few definitions as possible to obtain an environment in which to express theorems that require interesting proof techniques.
The theorems themselves are all well-known results, which suits our aim of exploring mechanisation techniques instead of new theory.
The small scale of the challenge problems also means that they are also accessible to mechanisation experts who may not be familiar with concurrency theory.
While the challenge problems themselves are thus not of scientific interest, we expect that the best practices and mechanisations developed by solving the challenge problems will also be useful for future research.

We have begun collecting solutions to our challenge problems on our website.
In the long term, we expect to use the website for promoting best practices, tutorials and guidelines derived from solutions to our challenge problems.
The interested reader can find more information about the current status at:
\begin{center}
  \url{https://concurrentbenchmark.github.io/}
\end{center}
We encourage anyone interested to try the challenge problems using their favorite tools and techniques, and to send us their solutions.

In the rest of this paper we discuss the overall design of the benchmark, describe our three challenge problems, and discuss possible extensions and further work.

@@

\section{The challenge problems}
In this section we will briefly describe our three challenge problems, referring to the appendix for the full text of each challenge.
Our \textbf{linearity challenge} (Appendix~\ref{sec:challenge:linearity-beh-types}) is to prove type preservation (also known as subject reduction) for a process calculus with a session type system.
Session type systems require linear handling of typing contexts, and the proof involves reasoning about the linearity of context splitting operations.
In name passing systems, scope extrusion is the expansion of the scope of a name when the name is sent outside of its original scope.
Our \textbf{scope extrusion challenge} (Appendix~\ref{sec:challenge:name-passing-scope-extrusion}) is to prove that barbed bisimulation is an equivalence relation for an untyped name passing calculus without infinite behaviour, but with non-deterministic choice.
The proof involves reasoning about names that are ``in the process'' of being scope extruded, which presents issues for some approaches to mechanising binders.
Our \textbf{coinduction challenge} (Appendix~\ref{sec:challenge:coinduction}) is to prove that strong barbed bisimulation is an equivalence relation \footnote{AM: this seems wrong, see:  \( \sbcong{P}{Q} \) if and only if for any process \( R \) and substitution \( \sigma \), \( \sbbisim{\Ppar{\applysubst{\sigma}{P}}{R}}{\Ppar{\applysubst{\sigma}{Q}}{R}} \).}for an untyped calculus of communicating systems with (infinite) replication of processes.
The proof involves reasoning coinductively about the infinite behaviours of the replication operator.
We invite anyone interested to send us solutions to the challenge problems and to publish experience reports detailing the advantages and disadvantages of their approaches.
